---
title: "In-class assignment 01 and 02: answer key"
subtitle: "Instructor copy. Do not put this on the course website."
format:
  html:
    embed-resources: true
    number-sections: true
    number-depth: 1
    theme: flatly
    toc: true
execute:
  echo: true
  message: false
  warning: false
---

Each question below shows an answer that earns full credit, then explains what
the question was actually testing and what to look for when grading.

# Q1: element types and what `c()` does to them

## (a) A numeric vector

```{r}
nums <- c(4, 8, 15, 16, 23, 42)
class(nums)
```

`class()` reports `"numeric"`, not `"integer"`. Numbers typed without the `L`
suffix are doubles, so `4` is a double and `4L` is an integer. Do not take
points off for a student who expected `"integer"`, but it is worth saying out
loud in class, because it is the first place the five element types stop being
an abstract list.

## (b) A vector holding a number and a word

```{r}
mixed <- c(1, 2, "three")
mixed
class(mixed)
```

## (c) What happened, and why

A vector holds one element type and one only. When `c()` is handed a mixture,
it does not fail and it does not keep both types. It converts everything to the
one type that can represent all the values, which here is character. That is
why `mixed` prints as `"1" "2" "three"` with quotation marks around the 1 and
the 2: they are no longer numbers, they are the text of numbers. The order R
follows is logical, then integer, then double, then character, always moving
toward the more accommodating type, never back.

`mixed[1] + 1` is therefore an error:

```{r}
#| error: true
mixed[1] + 1
```

R is telling you that `"1"` is text and addition has no meaning for text.

**What to look for.** The word "coercion" is not required, but the idea is: the
student has to say that R changed the numbers into text, and that it did so
because a vector cannot hold two types. An answer that says "R made an error"
or "the vector holds both types" has missed the point of the question.

## (d) Quarto part: hide the code

````{verbatim}
```{r}
#| echo: false
sum(nums)
```
````

```{r}
#| echo: false
sum(nums)
```

The result, 108, appears with no code above it. `#| echo: false` hides the code
and still runs it, which is exactly what you want in a report where the reader
cares about the number and not about how you got it.

**What to look for.** The chunk must still be in the `.qmd`, and the number
must still appear in the `.html`. Two common failures: deleting the code and
typing 108 by hand, and reaching for `#| include: false`, which hides the
result as well and leaves the reader with nothing.

# Q2: getting parts out of a `data.frame`

## (a) Build the data

```{r}
plots <- data.frame(
  plot_id = c("a", "b", "c", "d"),
  yield = c(180, 205, 172, 198),
  irrigated = c(TRUE, TRUE, FALSE, TRUE)
)

str(plots)
```

`str()` is the fastest way to see all three element types at once: `chr` for
the ids, `num` for the yields, `logi` for the irrigation flags. In older
versions of R the `plot_id` column would have come back as a factor rather than
character. Since R 4.0.0 it does not, so a student on a current install sees
`chr`.

## (b) Three ways in

```{r}
plots$yield

plots[["yield"]]

plots[3, ]
```

The first two return the same thing, a plain numeric vector of four values.
`$` is the convenient form you will type most often. `[[ ]]` does the same job
but takes the column name as a string, which is what lets you hold the name in
a variable and pull a column whose name you do not know until the code runs.

`plots[3, ]` is different in kind. The comma splits rows from columns, rows
before it and columns after, and leaving the column side empty means "all
columns". So it returns a one-row `data.frame`, not a vector: row three, every
column.

**What to look for.** Both column methods present, and the row pulled with a
comma. A student who writes `plots[3]` has selected the third **column**, which
is a different answer, and it is worth marking, because that comma is the piece
that comes back in every later chapter.

## (c) A `data.frame` is a list with a rule

Every column of a `data.frame` must have the same number of entries, because
the columns line up into rows. An ordinary list has no such rule: its elements
can be any length, and any type, including other lists. Both are lists of
elements where each element can be a different type. Only the `data.frame`
insists that those elements be equal in length, and that is what makes rows
meaningful.

## (d) Quarto part: inline code

The sentence in the assignment, written with inline R code:

````{verbatim}
The `plots` data has `r nrow(plots)` rows and a mean yield of `r mean(plots$yield)` bushels.
````

which renders as:

> The `plots` data has `r nrow(plots)` rows and a mean yield of
> `r mean(plots$yield)` bushels.

**What to look for.** Read the `.qmd`, not the `.html`. Both numbers rendered
correctly tells you nothing on its own, since a student who typed 4 and 188.75
by hand gets an `.html` that looks identical. The credit is for the backtick-r
form in the source. This is the habit worth building: a number typed by hand is
a number that goes stale the moment the data changes, and nobody notices.

# Q3: functions, arguments, and packages

## (a) `seq()` with named arguments

```{r}
steps <- seq(from = 0, to = 100, by = 5)
length(steps)
```

21, not 20, because both ends are included. Naming the arguments is the point
of the question. `seq(0, 100, 5)` returns the same vector, but the reader has
to know `seq()`'s argument order to see what it does, and you have to remember
it to write it. `from =`, `to =`, `by =` says it out loud.

## (b) `NA` is contagious

```{r}
steps_na <- c(steps, NA)
mean(steps_na)

mean(steps_na, na.rm = TRUE)
```

The first call returns `NA`. That is not a failure, it is the honest answer:
one of the values is unknown, so the average of the values is unknown too.
The argument `na.rm = TRUE` tells `mean()` to drop the missing values first and
average what is left, which gives 50.

**What to look for.** The student names the argument `na.rm` and says what it
does. The deeper point, worth making in class, is that R will not silently skip
missing values for you. Being told `NA` is how you find out your data has holes
in it.

## (c) Reading the error message

```{r}
#| error: true
mean(steps, na.rm)
```

Written without `= TRUE`, `na.rm` is not an argument name any more. It is an
ordinary R expression, so R goes looking for an object called `na.rm`, finds
nothing by that name, and stops. The message names the thing it could not find,
which is the useful habit here: read the message, find the name in it, and ask
why R expected that name to exist.

**What to look for.** The error quoted, and an explanation that R treated
`na.rm` as an object rather than as an argument name. "It needs to be
`na.rm = TRUE`" alone is the fix, not the reason, so give partial credit.

## (d) Quarto part: shown but not run, and quiet loading

````{verbatim}
```{r}
#| eval: false
install.packages("dplyr")
```
````

```{r}
#| eval: false
install.packages("dplyr")
```

`#| eval: false` shows the code and never runs it. Installing a package is a
once-per-machine job, so a reader should see the command, but the document
should not run it on every render, which would be slow and would reinstall a
package that is already there.

````{verbatim}
```{r}
#| message: false
#| warning: false
library(dplyr)
glimpse(plots)
```
````

```{r}
#| message: false
#| warning: false
library(dplyr)
glimpse(plots)
```

`library()` prints startup chatter, in the case of dplyr a note about masked
functions. It is useful in the console and it is noise in a report, so
`#| message: false` and `#| warning: false` keep it out.

**What to look for.** `eval: false` on the install chunk and both options on the
library chunk. Watch for students who set the two options on the wrong chunk, or
who use `#| echo: false` on the install chunk, which hides the very command the
reader is supposed to see.

Note that this document sets `message: false` and `warning: false` for every
chunk in its YAML header, under `execute:`, which is how you would normally
silence a whole report rather than repeating the option chunk by chunk. The
question asks for the per-chunk form because that is the one worth learning
first.
