```{r}
#| include: false
library(lubridate)
library(dplyr)
library(data.table)
library(flextable)
```

# Instructor:

Taro Mieno:

+ Email: tmieno2@nebraska.edu
+ Teaching Assistant: none

# Schedule

+ Lectures:
  - Monday, Tuesday, Wednesday: 1限 - 4限
  - Thursday: 1限 - 3限

# Course Description:

The goal of this course is to prepare students for jobs that require quantitative skills beyond Microsoft Excel and graduate programs. The R software is used throughout the course. In order to achieve the goal, students will be introduced to the basics of programming and how to apply it to real world issues in the field of agricultural (agricultural economics, agronomy, etc) and environmental sciences. By completing the course, students will know data wrangling (e.g., merging, transforming datasets), data visualization, and exploratory data analysis, spatial data management.

# Reading Materials

+ Recommended:

  - Grolemund, Garrett. and Wickham, Hadley. 2019 \href{https://r4ds.had.co.nz/}{"R for Data Science"}
  - Lovelace, Robin., Nowosad, Jakub., and Muenchow, Jannes. 2019 \href{https://geocompr.robinlovelace.net/}{"Geocomputation with R"}

+ Prerequisites:

  - Introductory statistics or equivalent

# Grading

```{r}
#| echo: false
grade_table <-
  data.table(
    Assignment = c(
      "In-class assignment 01 and 02",
      "In-class assignment 03-2",
      "In-class assignment 03-3",
      "In-class assignment 04-1",
      "In-class assignment 04-2",
      "In-class assignment 04-3",
      "In-class assignment 05-1",
      "Total"
    ),
    Points = c("10", "15", "15", "15", "15", "15", "15", "100")
  )

flextable(grade_table) %>% autofit()
```

+ **In-class assignments**: There will be 7 in-class assignments, one for each lecture note. The first is worth 10 points and the other six are worth 15 points each, for 100 points in total. Each is worked on and submitted during class.

# Topics Covered

+ Introduction to R
+ Quarto
+ Data Wrangling
+ Merge and reshape datasets
+ Data visualization
+ Writing your own function
+ Looping
+ Parallel computing
+ Writing reproducible articles

# Academic Honesty:

Students are expected to follow Kobe University's regulations on academic honesty. The work you hand in must be your own. Students are encouraged to contact the instructor for clarification of these guidelines if they have questions or concerns.

# Students with disabilities:

Students with disabilities are encouraged to contact the instructor for a confidential discussion of their individual needs for academic accommodation. Accommodations that may affect a student's ability to participate fully in course activities, or to meet course requirements, are arranged through Kobe University's student support services, and students who need them should contact that office as well as the instructor.
