```{r}
#| include: false
library(lubridate)
library(dplyr)
library(data.table)
library(flextable)
```

# Instructor:

Taro Mieno:

+ Email: tmieno2@nebraska.edu
+ Teaching Assistant: none

# Schedule

+ Lectures:
  - Monday, Tuesday, Wednesday: 1限 - 4限
  - Thursday: 1限 - 3限

# Course Description:

The goal of this course is to prepare students for jobs that require quantitative skills beyond Microsoft Excel and graduate programs. The R software is used throughout the course. In order to achieve the goal, students will be introduced to the basics of programming and how to apply it to real world issues in the field of agricultural (agricultural economics, agronomy, etc) and environmental sciences. By completing the course, students will know data wrangling (e.g., merging, transforming datasets), data visualization, and exploratory data analysis, spatial data management.

# Reading Materials

+ Recommended:

  - Grolemund, Garrett. and Wickham, Hadley. 2019 \href{https://r4ds.had.co.nz/}{"R for Data Science"}
  - Lovelace, Robin., Nowosad, Jakub., and Muenchow, Jannes. 2019 \href{https://geocompr.robinlovelace.net/}{"Geocomputation with R"}

+ Prerequisites:

  - Introductory statistics (STAT 218) or equivalent

# Grading

```{r}
#| echo: false
grade_table <-
  data.table(
    Assignment = c(
      "In-class assignment 01 and 02",
      "In-class assignment 03-2",
      "In-class assignment 03-3",
      "In-class assignment 04-1",
      "In-class assignment 04-2",
      "In-class assignment 04-3",
      "In-class assignment 05-1",
      "Total"
    ),
    Points = c("10", "15", "15", "15", "15", "15", "15", "100")
  )

flextable(grade_table) %>% autofit()
```

+ **In-class assignments**: There will be 7 in-class assignments, one for each lecture note. The first is worth 10 points and the other six are worth 15 points each, for 100 points in total. Each is worked on and submitted during class.

# Topics Covered

+ Introduction to R
+ Quarto
+ Data Wrangling
+ Merge and reshape datasets
+ Data visualization
+ Writing your own function
+ Looping
+ Parallel computing
+ Writing reproducible articles

# Academic Honesty:

Students are expected to adhere to guidelines concerning academic dishonesty outlined in [Section 4.2 of University's Student Code of Conduct](\url{http://stuafs.unl.edu/ja/code/}). Students are encouraged to contact the instructor for clarification of these guidelines if they have questions or concerns. The Department of Agricultural Economics has a written policy defining academic dishonesty, the potential sanctions for incidents of academic dishonesty, and the appeal process for students facing potential sanctions. The Department also has a policy regarding potential appeals of final course grades. These policies are available for review on [the department’s website](\url{http://agecon.unl.edu/undergraduate})

# Students with disabilities:

Students with disabilities are encouraged to contact the instructor for a confidential discussion of their individual needs for academic accommodation. It is the policy of the University of Nebraska-Lincoln to provide flexible and individualized accommodation to students with documented disabilities that may affect their ability to fully participate in course activities or to meet course requirements. To receive accommodation services, students must be registered with the Services for Students with Disabilities (SSD) office, 132 Canfield Administration, 472-3787 voice or TTY.
